---
title: "Data Wrangling"
author: "Erin Anthony"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Read in the data
```{r read in the data}
eduData = read.csv("csv-data/all-data.csv") # 7357 obs of 150 variables
```

### Examine the variable levels of the data selection variables
```{r examine the variable levels of the data selection variables}
levels(eduData$HD2015.Undergraduate.offering) # remove "No undergraduate offering"
levels(eduData$HD2015.Degree.granting.status) # remove "Nondegree-granting, primarily postsecondary"
levels(eduData$HD2015.Institution.is.active.in.current.year) # remove missing values
levels(eduData$IC2015.Bachelor.s.degree) # remove missing values, remove "Implied no", remove "Not applicable"
```

### Examine the variable levels of the independent variables
```{r examine the variable levels of the independent variables}
levels(eduData$HD2015.Bureau.of.Economic.Analysis..BEA..regions) # categorical, missing values
levels(eduData$HD2015.Historically.Black.College.or.University) # binary, missing values
levels(eduData$HD2015.Tribal.college) # binary, missing values
levels(eduData$HD2015.Degree.of.urbanization..Urban.centric.locale.) # categorical, missing values
levels(eduData$IC2015.Open.admission.policy) # binary, missing values, remove "Not applicable"
levels(eduData$HD2015.Land.Grant.Institution) # binary, remove missing values
levels(eduData$IC2015.Associate.s.degree) # binary, missing values
levels(eduData$IC2015.All.programs.offered.completely.via.distance.education) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Undergraduate.programs.or.courses.are.offered.via.distance.education) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Study.abroad) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Weekend.evening..college) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Remedial.services) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Academic.career.counseling.service) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Employment.services.for.students) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Placement.services.for.completers) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.On.campus.day.care.for.students..children) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Tuition.guaranteed.plan) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Prepaid.tuition.plan) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Tuition.payment.plan) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Institution.provide.on.campus.housing) # binary, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015.Institution.provides.board.or.meal.plan) # binary, two "Yes" levels, missing values, "Not applicable", "Not reported"
levels(eduData$IC2015_AY.In.state.average.tuition.for.full.time.undergraduates) # ratio, missing values
levels(eduData$IC2015_AY.Out.of.state.average.tuition.for.full.time.undergraduates) # ratio, missing values
levels(eduData$HD2015.Institution.size.category) # categorical, missing values, "Not applicable", "Not reported"
levels(eduData$EF2015D.Student.to.faculty.ratio) # ratio, missing values
levels(eduData$SFA1415.Average.amount.of.federal..state..local..institutional.or.other.sources.of.grant.aid.awarded.to.undergraduate.students) # ratio, missing values
levels(eduData$SFA1415.Average.amount.of.federal.student.loans.awarded.to.undergraduate.students) # ratio, missing values
```

### Examine the variable levels of the dependent variables
```{r examine the variable levels of the dependent variables}
levels(eduData$DRVEF2015.Percent.of.undergraduate.enrollment.that.are.Black.or.African.American) # ratio, missing values
levels(eduData$DRVEF2015.Percent.of.undergraduate.enrollment.that.are.Hispanic.Latino) # ratio, missing values
```

### Trim the data set and rename the columns
```{r trim the data set and rename the columns}
edu.df <- eduData[, c(10,12,18,26,8,13,14,15,17,21,24,34,35,38,39,40,
                      41,42,43,44,46,47,48,49,50,51,53,71,113,117,121,
                      93,94)] # 7357 obs of 33 variables

colnames(edu.df) <- c("ugrad", "degree", "active", "bachelors", "region", "hbcu", "tribal", "urban", "open", "landGrant", "associates", "allDistance", "partDistance", "studyAbroad", "weekend", "remedial", "counseling", "employment", "placement", "dayCare", "tuitionGuarantee", "tuitionPrepaid", "tuitionPlan", "oncampusHousing", "mealPlan", "instateTuition", "outstateTuition", "size", "facultyRatio", "grantAid", "loanAid", "blackEnroll", "latinoEnroll")
```

### Only include active institutions granting Bachelor\'s degrees
```{r only include active institutions granting Bachelor\'s degrees}
edu.df <- subset(edu.df, ugrad=="Undergraduate degree or certificate offering" & degree=="Degree-granting" & active=="Yes" & bachelors=="Yes") # 2752 obs of 33 variables
edu.df$ugrad <- NULL
edu.df$degree <- NULL
edu.df$active <- NULL
edu.df$bachelors <- NULL # 29 variables
```

### Combine black and latino enrollments to use as the target variable
```{r combine black/latino enrollments to use as the target variable}
edu.df$minorityEnroll <- edu.df$blackEnroll + edu.df$latinoEnroll
edu.df$blackEnroll <- NULL
edu.df$latinoEnroll <- NULL # 28 variables
```

### Examine the data to determine missing value imputation methods and feature engineer some new independent variables
```{r examine the data to determine missing value imputation methods}
# remove rows where data is missing from the dependent variable column
edu.df <- edu.df[!is.na(edu.df$minorityEnroll),] # 2698 obs of 28 var

# remove outliers from the minority enrollment data (top/bottom 2.5%)
quantile(edu.df$minorityEnroll, 0.975) # 91%
quantile(edu.df$minorityEnroll, 0.025) # 0%
edu.df <- edu.df[edu.df$minorityEnroll < quantile(edu.df$minorityEnroll, 0.975), ] # 2629 obs of 28 variables

# find columns that still contain missing values
colSums(is.na(edu.df)) # instateTuition (39), outstateTuition (39), grantAid (17), loanAid (123)

table(edu.df$region) # 9 levels, categorical variable
# replace levels with shorter name codes
edu.df$region <- with(edu.df, ifelse(region=="Far West AK CA HI NV OR WA", "FW", ifelse(region=="Great Lakes IL IN MI OH WI", "GL", ifelse(region=="Mid East DE DC MD NJ NY PA", "ME", ifelse(region=="New England CT ME MA NH RI VT", "NE", ifelse(region=="Plains IA KS MN MO NE ND SD", "PL", ifelse(region=="Rocky Mountains CO ID MT UT WY", "RM", ifelse(region=="Southeast AL AR FL GA KY LA MS NC SC TN VA WV", "SE", ifelse(region=="Southwest AZ NM OK TX", "SW", "US")))))))))
table(edu.df$region)

table(edu.df$hbcu) # 2586 No, 43 Yes
table(edu.df$tribal) # 2616 No, 13 Yes

table(edu.df$urban) # 12 levels, categorical variable
# replace levels with shorter, ordered names
edu.df$urban <- with(edu.df, ifelse(urban=="City: Large", "City:L", ifelse(urban=="City: Midsize", "City:M", ifelse(urban=="City: Small", "City:S", ifelse(urban=="Rural: Distant", "Rural:D", ifelse(urban=="Rural: Fringe", "Rural:F", ifelse(urban=="Rural: Remote", "Rural:R", ifelse(urban=="Suburb: Large", "Suburb:L", ifelse(urban=="Suburb: Midsize", "Suburb:M", ifelse(urban=="Suburb: Small", "Suburb:S", ifelse(urban=="Town: Distant", "Town:D", ifelse(urban=="Town: Fringe", "Town:F", "Town:R"))))))))))))
edu.df$urban <- ordered(edu.df$urban, levels = c("Rural:R", "Rural:D", "Rural:F", "Town:R", "Town:D", "Town:F", "Suburb:S", "Suburb:M", "Suburb:L", "City:S", "City:M", "City:L"))
table(edu.df$urban)

table(edu.df$open) # 1836 No, 663 Yes, 130 Not applicable
# replace 'N/A' values in the 'open' column with 'No' values
edu.df$open[edu.df$open=="Not applicable"] <- "No"
table(edu.df$open) # 1966 No, 663 Yes

table(edu.df$landGrant) # 2552 Not a Land Grant, 77 Land Grant
# replace levels with No and Yes
edu.df$landGrant <- with(edu.df, ifelse(landGrant=="Land Grant Institution", "Yes", "No"))
table(edu.df$landGrant)

table(edu.df$associates) # 1157 No, 1472 Yes
table(edu.df$allDistance) # 2590 No, 39 Yes
table(edu.df$partDistance) # 1972 Yes, 657 No
table(edu.df$studyAbroad) # 1543 Yes, 1086 No
table(edu.df$weekend) # 1343 No, 1286 Yes
table(edu.df$remedial) # 1818 Yes, 811 No
table(edu.df$counseling) # 2583 Yes, 46 No

table(edu.df$employment) # 2245 Yes, 384 No
table(edu.df$placement) # 2083 Yes, 546 No
# combine employment and placement services into a single variable
edu.df$placeEmploy <- with(edu.df, ifelse(employment=="Yes" | placement=="Yes", "Yes", "No"))
edu.df$employment <- NULL
edu.df$placement <- NULL # 27 variables
table(edu.df$placeEmploy) # 2326 Yes, 303 No

table(edu.df$dayCare) # 2167 No, 462 Yes

table(edu.df$tuitionGuarantee) # 2467 No, 162 Yes
table(edu.df$tuitionPrepaid) # 2371 No, 258 Yes
table(edu.df$tuitionPlan) # 2195 Yes, 434 No
# combine tuition guarantee, prepaid tuition, and tuition payment plan
edu.df$tuitionOptions <- with(edu.df, ifelse(tuitionGuarantee=="Yes" | tuitionPlan=="Yes" | tuitionPrepaid=="Yes", "Yes", "No"))
edu.df$tuitionGuarantee <- NULL
edu.df$tuitionPrepaid <- NULL
edu.df$tuitionPlan <- NULL # 25 variables
table(edu.df$tuitionOptions) # 2234 Yes, 395 No

table(edu.df$oncampusHousing) # 1717 Yes, 912 No

table(edu.df$mealPlan) # 1574 Yes (2 levels), 1055 No
# combine the two 'Yes' levels into one level
edu.df$mealPlan <- with(edu.df, ifelse(mealPlan=="Yes, number of meals in the maximum meal plan offered" | mealPlan=="Yes, number of meals per week can vary", "Yes", "No"))
table(edu.df$mealPlan) # 1574 Yes, 1055 No

# table(edu.df$instateTuition) # numerical, 39 missing values
# examine the average and spread of in-state tuition
mean(edu.df$instateTuition, na.rm=TRUE) # 18335
quantile(edu.df$instateTuition, na.rm=TRUE) # 0, 8953, 15568, 26240, 55944
# replace N/A values in the 'instateTuition' column with the mean
edu.df$instateTuition[is.na(edu.df$instateTuition)] <- 18335
colSums(is.na(edu.df))

# remove outliers from the in-state tuition data (top/bottom 2.5%)
quantile(edu.df$instateTuition, 0.975) # 47150
quantile(edu.df$instateTuition, 0.025) # 2849
edu.df <- edu.df[edu.df$instateTuition < quantile(edu.df$instateTuition, 0.975) & edu.df$instateTuition > quantile(edu.df$instateTuition, 0.025), ] # 2497 obs of 25 variables

# table(edu.df$outstateTuition) # numerical, 39 missing values
# calculate all numerical variable correlations for tuition
cor(edu.df[, "outstateTuition"], edu.df[, sapply(edu.df, is.numeric)], use="complete.obs")
# there is an 87% correlation between out-of-state and in-state tuition
# remove out-of-state tuition from the analysis
edu.df$outstateTuition <- NULL # 24 variables

table(edu.df$size) # 5 levels, categorical variable
# replace levels with shorter, ordered names
edu.df$size <- with(edu.df, ifelse(size=="Under 1,000", "<1,000", ifelse(size=="1,000 - 4,999", "1-4,999", ifelse(size=="5,000 - 9,999", "5-9,999", ifelse(size=="10,000 - 19,999", "10-19,999", "20,000+")))))
edu.df$size <- ordered(edu.df$size, levels=c("<1,000", "1-4,999", "5-9,999", "10-19,999", "20,000+"))
table(edu.df$size)

# table(edu.df$facultyRatio) # numerical, 0 missing values
# remove outliers from the student-faculty ratio data (top/bottom 2.5%)
quantile(edu.df$facultyRatio, 0.975) # 26%
quantile(edu.df$facultyRatio, 0.025) # 5%
edu.df <- edu.df[edu.df$facultyRatio < quantile(edu.df$facultyRatio, 0.975) & edu.df$facultyRatio > quantile(edu.df$facultyRatio, 0.025), ] # 2338 obs of 24 variables

# table(edu.df$grantAid) # numerical, 17 missing values
# calculate all numerical variable correlations for grant aid
cor(edu.df[, "grantAid"], edu.df[, sapply(edu.df, is.numeric)], use="complete.obs")
# there is a correlation of 80% between grant aid and in-state tuition
# create a linear regression model to predict grant aid
grant.mod <- lm(grantAid ~ instateTuition, data=edu.df)
summary(grant.mod)
# use the model to impute missing values
grant.na <- edu.df[is.na(edu.df$grantAid),]
grant.vals <- data.frame(predict(grant.mod, newdata=grant.na))
colnames(grant.vals) <- "grantAid"
for(n in row.names(grant.vals)) {
  edu.df[n,"grantAid"] <- grant.vals[n,"grantAid"]
}
colSums(is.na(edu.df))

# remove outliers from the grant aid data (top/bottom 2.5%)
quantile(edu.df$grantAid, 0.975) # 26516
quantile(edu.df$grantAid, 0.025) # 2580
edu.df <- edu.df[edu.df$grantAid < quantile(edu.df$grantAid, 0.975) & edu.df$grantAid > quantile(edu.df$grantAid, 0.025), ] # 2220 obs of 24 variables

# table(edu.df$loanAid) # numerical, 123 missing values
# calculate all numerical variable correlations for loan aid
cor(edu.df[, "loanAid"], edu.df[, sapply(edu.df, is.numeric)], use="complete.obs") # there are no strong correlations
# examine the average and spread of loan aid
mean(edu.df$loanAid, na.rm=TRUE) # 7554
quantile(edu.df$loanAid, na.rm=TRUE) # 1154 6771 7358 88305 14400
# replace N/A values in the 'loanAid' column with the mean
edu.df$loanAid[is.na(edu.df$loanAid)] <- 7554
colSums(is.na(edu.df)) # no missing values :) 
```

### Write the data to a new csv file
```{r write the data to a new csv file}
write.csv(edu.df, file = "project-data.csv", row.names=FALSE)
```

### Data wrangling summary

I began with a cursory examination of the data. I next created a new data frame containing only the variables that I plan to use in my analysis and renamed them with shorter names that will be easier to use in my code. I only wish to examine data from institutions that are actively granting Bachelor's degrees, so I eliminated the data from any institutions that did not meet these criteria. I also decided to combine the enrollment numbers for African American and Hispanic/Latino students into one variable to use as my target variable. 

I next conducted a more in-depth examination of my selected data set to determine missing value imputation methods and feature engineer some new independent variables. I first removed all rows that did not contain any data for my target variable, and then examined the remaining data for additional missing values. I found that four of my independent variables still contained missing values: instateTuition (40 missing), outstateTuition (40 missing), grantAid (18 missing), and loanAid (127 missing).

I found a very high (90%) correlation between in-state and out-of-state tuition, so I removed out-of-state tuition from the data set and replaced the missing values for in-state tuition with the mean value for in-state tuition. I also noticed a strong (82%) correlation between grant aid and in-state tuition, and I decided to create a linear regression model to predict grant aid based on in-state tuition. I then used the model to impute new values and replaced the missing values with the imputed ones. I did not find any strong correlations for loan aid and replaced the missing values with the mean value.

I noticed that the region, urban, and size variable level names were quite long, so I shortened them to work better in the data visualizations, and ordered the size and urban levels in ascending order by size and degree of urbanization, respectively. I replaced the 'not applicable' values of the open enrollment variable with 'no' values, since this did imply that open enrollment was not in play. 

I combined the availability of employment services and the availability of placement services into a single variable, since these services are very similar in nature. I also combined tuition guarantee, tuition prepaid, and tuition plan into a single variable to indicate the availability of some type of special tuition payment option.

I also removed the top and bottom 2.5% of the numerical data to get rid of outliers.